---
title: 'Object Oriented Programming in R'
author: 'André Victor Ribeiro Amaral'
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Introduction

In this tutorial, we will introduce the Object Oriented Programming (OOP) paradigm in `R`. Although, `R` can be seen more as a [Functional Language](https://en.wikipedia.org/wiki/Functional_programming), different OOP systems can be used. In particular, we will cover `S3` and `R6` (as we will define below). The main reference here is the [`Advanced R` book](https://adv-r.hadley.nz/index.html).

---

Different from other programming languages, there are (at least) 4 different OOP systems available in `R`, namely, `S3`, `S4`, `RC` (which stands for Reference Class), and `R6` (from the `R6` package). Also, although OOP is important in `R`, `R` can be seen as an Functional Language, that is, a language made of *first-class functions* (functions that behave like any other data structure) that are used to solve problems.

In this regard, we also have to make the distinction between **encapsulated OOP** and **functional OOP**. The former works like in most of the other languages (e.g., Python, Java, etc.), in which the methods belong to objects or classes, and *method calls* typically look like `object.method(arg)`, and the latter works through generic functions, and *method calls* look like ordinary function calls, as in `generic(arg)`.

In `R`,

- `S3` is an informal implementation of **functional OOP** and relies on common conventions. Functions have different behaviors depending on the arguments’ class.
- `S4` is a formal and rigorous rewrite of `S3`. It is also an implementation of **functional OOP**.
- `RC` implements **encapsulated OOP**.
- `R6` implements **encapsulated OOP**, like `RC`, but resolves some issues.

<div class = 'space'></div>

# OOP

In Object Oriented Programming (OOP) paradigm (more specifically, in **encapsulated OOP**), we have (among other concepts) **classes**, **objects**, **methods**, and **attributes**. Roughly speaking, a `class` is a definition of an object, an `object` is an instance of a class created with specifically defined data, a `method` is a function that describes the behavior of an object, and an `attribute` is a variable associated with an object that represents its state.

However, In **base** `R`, things are a bit different. Although one may see online that “*everything that exists in `R` is an object*”, we have to distinct **base objects** from **OO objects**. OO objects have a class attribute. For instance,
```{r}
ch <- character()
is.object(x = ch)
df <- as.data.frame(matrix(data = rnorm(n = 4), nrow = 2, ncol = 2))
is.object(x = df)
attr(ch, 'class')
attr(df, 'class')
```

Every object has a base type, though.  You can access it with the `typeof()` function.
```{r}
typeof(ch)
typeof(df)
```

<div class = 'space'></div>

## `S3`

`S3` is the simplest OOP implemented system. As it is a **functional OOP** implementation, it will not follow the commonly used notation for creating objects and calling methods (e.g., `object.method(arg)`); instead we will mostly use them along with **polymorphic** functions (i.e., functions that behave differently depending on their input).

Let's start with the following example
```{r}
fact <- factor(x = c(1, 2, 1, 1, 2), levels = c(1, 2), labels = c('first', 'second'))
fact
is.object(x = fact)
attr(fact, 'class')
typeof(x = fact)
attributes(fact)
```

Notice that `fact` is an OO object with class `factor`. In this regard, we can get the underlying base type of `fact` by using the `unclass()` function.
```{r}
unclass(fact)
```

A (not so) easy way to test if an object is a `S3` object is doing the following (i.e., testing if it is a OO object and if it is **not** a `S4` object)
```{r}
is.object(fact) & (!isS4(fact))
```

<div class = 'space'></div>

### Generic functions 

However, the main advantage of using such a system is related to the **generic functions**. In that case, depending on the input class, the functions behaves differently. For instance, let's see how the `summary()` function works with `fact` and with the data frame `df`.
```{r}
summary(fact)
summary(df)
```

For the `fact` object, `summary()` returns the count of observations in each class, and for the `df` object, it returns a few statistics that can be computed for each numeric vector (represented by `df`'s columns).

If we want to check if a particular function is a **generic function**, we can use the `ftype()` function from the `sloop` package.
```{r}
library(sloop)
sloop::ftype(f = summary)
```

Also, if we want to see how `summary(fact)` is different from `summary(nv)`, we can call `summary.factor` and `summary.data.frame`.
```{r}
summary.factor
summary.data.frame
```

<div class = 'space'></div>

### Creating new classes

Now, let's create a new class and make an object that is an instance of such a class. This can be done using the `structure()` function, or in a two-step procedure
```{r}
# One step
obj <- structure('André', class = 'person')
# Two steps
obj <- 'André'
class(obj) <- 'person'
attr(obj, 'class')

is.object(obj)
```

However, because there is no validation on the how a class can be assigned to an object, when creating a new class, it is better to provide at least three functions: a *constructor*, a *validator*, and a *helper*.

```{r, error = TRUE}
# Constructor
new_person <- function (x = character()) {
  if (!is.character(x)) { stop('"x" should be a "character".') }
  structure(x, class = 'person')
}

obj <- new_person(1)
obj <- new_person('André')
attr(obj, 'class')

# Validator
validate_person <- function (x) {
  values <- unclass(x)
  if (length(values) == 0) { stop('Object of this class must be have length >= 1.') }
  x
}
obj <- validate_person(new_person(character()))
obj <- validate_person(new_person('André'))
print.default(obj)


# Helper (coerce the input to a valid value)
person <- function (x) { # same name as the class
  if (length(x) == 0) { x <- 'André' }
  validate_person(new_person(x))
}
obj <- person(character())
print.default(obj)
```
Since we have a new class, we can also code new behaviors for any **generic function** we want. For instance, we can customize the `print()` function for an object of class `person`. To do so, we can use the notation `generic_function.class_name()` when creating the functions; in particular
```{r}
print.person <- function (x, ...) {
  x <- unclass(x)
  m <- paste('Hi, my name is ', x, '.', sep = '')
  print(m)
  invisible(m)
}

print(obj)
```

To obtain a list of the possible choices for `arg` in `print(arg)`, one can use the `sloop::s3_methods_generic()` function.
```{r}
sloop::s3_methods_generic('print')$class
```


<div class = 'space'></div>

### Creating new generic functions

As the `summary()` and `print()` functions, we can also create our own **generic functions**. To do so, we have to use the `UseMethod()` function. For example, we will create the `display()` function that will "mimic" the `print()` function output.
```{r}
display <- function (x, ...) {
  UseMethod('display')
}
```

Now, all we have to do is creating the specific behaviors depending on the object class.
```{r}
display.person <- function (x, ...) {
  x <- unclass(x)
  m <- paste(x, ' is a person.', sep = '')
  print(m)
  invisible(m)
}

display.animal <- function (x, ...) {
  x <- unclass(x)
  m <- paste(x, ' is an animal', sep = '')
  print(m)
  invisible(m)
}
```

We can create objects from the `person` and `animal` (and others) classes as before.
```{r}
p <- person('André')
a <- structure('Dog', class = 'animal')
o <- structure('Key', class = 'object')
```

And then, using the `display()` function for each of these objects.
```{r, error = TRUE}
display(p)
display(a)
display(o)
```

In the above example, notice that, since we did not create a function `display.object()`, the `display()` function did not any coded behavior for an object with class `object`. To overcome this issue, we can create a `default` behavior for `display()`.
```{r}
display.default <- function (x, ...) {
  x <- unclass(x)
  m <- paste(x, ' is not a person or an animal.', sep = '')
  print(m)
  invisible(m)
}

display(o)
```

<div class = 'space'></div>

More about `S3` can be seen [here](https://adv-r.hadley.nz/s3.html).

## `R6` 

`R6` uses the **encapsulated OOP** paradigm, and therefore, it looks more like OOP in other languages. In particular, we will use the `object$method(arg)` notation, which is similar to the `object.method(arg)` notation from, for example, Python.

As in the Introduction, to use the `R6` system, we have to install and load the `R6` library.
```{r}
# install.packages('R6')
library(R6)
```

<div class = 'space'></div>

### Creating new classes

And from the `R6` package, we will be only interested in the `R6Class()` function to create classes and their methods. To do this, we can proceed as follows
```{r}
Students <- R6Class(classname = 'Students', public = list(
  n_students = 0,
  ids = c(),
  names = c(), 
  add_student = function (id, name, ...) {
    if ((length(id) > 1) | (length(name) > 1)) { stop('Only one student should be added each time.') }
    self$n_students <- self$n_students + 1
    self$ids <- c(self$ids, id)
    self$names <- c(self$names, name)
    invisible(self)
  }
)) 
```
A few things to be noticed from the above code:

- The class name (`Students`) is the same as the `classname` parameter (`Students`).
- For the `public` parameter (`private` and `active` also exist), we are defining **attributes** (`n_students`, `ids`, `names`) and **methods** (`add_student()`).
- Inside the `add_student()` method, we have to use the `self$attribute` notation to refer to the objects' attributes.
- After calling a method, we are returning the object itself, so that we can use a chained notation (as we will see below).

Then, we can print the class `Students`
```{r}
Students
```

To create an object from the `Students` class, all we have to do is using the `ClassNAme$new()` notation
```{r}
STAT294 <- Students$new()
STAT294
```

After creating an object, we can access its attributes and methods in the following way
```{r}
STAT294$n_students
STAT294$ids
STAT294$names

STAT294$add_student(id = 179600, name = 'ANDRE')

# We can also see the object attributes by printing it.
STAT294
```

Also, as we returned `self` in the `add_students()` method, we can do the following
```{r, error = TRUE}
STAT294$add_student(id = 111111, name = 'MARIA')$add_student(id = 222222, name = 'JOHN')

# We can also see the object attributes by printing it.
STAT294
```

From the above, notice that now we have 3 students (also including `ANDRE`) in the `STAT294` object, meaning that changes in `R6` objects are preserved (without having to assign it to other variable, as we usually do in `R`).

<div class = 'space'></div>

### Important methods

<div class = 'space'></div>

As in the [`Advanced R` book](https://adv-r.hadley.nz/r6.html#r6-important-methods), there are two important methods that we might want to use when creating a class in `R6`, namely `initialize()` and `print()`. `initialize()` is executed when an object is created, and `print()` is executed when we try to print the object. Let's see how they work.
```{r}
Students <- R6Class(classname = 'Students', public = list(
  n_students = 0,
  ids = c(),
  names = c(), 
  initialize = function (ids, names, ...) {
    if (length(ids) != length(names)) { stop('length(ids) != length(names)') }
    self$n_students <- self$n_students + length(ids)
    self$ids <- c(self$ids, ids)
    self$names <- c(self$names, names)
  },
  print = function (...) {
    cat('Number of students: ', self$n_students, '. \n', sep = '')
    cat('IDs: ', paste(self$ids, collapse = ', '), '.\n', sep = '')
    cat('Students: ', paste(self$names, collapse = ', '), '.\n', sep = '')
    invisible(self)
  },
  add_student = function (id, name, ...) {
    if ((length(id) > 1) | (length(name) > 1)) { stop('Only one student should be added each time.') }
    self$n_students <- self$n_students + 1
    self$ids <- c(self$ids, id)
    self$names <- c(self$names, name)
    invisible(self)
  }
)) 
```

Now, we can create another object and add students at the same time. And when printing the object, the `print()` method from the `Students` class will be called.
```{r}
STAT394 <- Students$new(ids = c(123456, 654321), names = c('PAULA', 'JONATHAN'))
STAT394
```

<div class = 'space'></div>

### Inheritance

If we want to inherit the behavior from a parent class (e.g., `Students`), all we have to do is creating a child class with the parameter `inherit = ParentClass`, as follows
```{r}
StudentsV2 <- R6Class(classname = 'StudentsV2', inherit = Students, public = list(
  remove_student = function (id) {
    if (length(id) > 1) { stop('Only one student can be removed each time.') }
    id_tbd <- which(id %in% self$ids)
    self$ids <- self$ids[-id_tbd]
    self$names <- self$names[-id_tbd]
    
    invisible(self)
  }
))

STAT494 <- StudentsV2$new(ids = c(123123, 321321), names = c('JAMES', 'MARY'))
STAT494
STAT494$remove_student(id = 123123)
STAT494
```

Notice that, although we created an object based on the `StudentsV2` class, we still have access to all attributes and methods from the parent class (`Students`). Also, we extended it by including the `remove_students()` method.

More about `R6` can be seen [here](https://adv-r.hadley.nz/r6.html).

<div class = 'space'></div>

# Example using `S3`

Aiming to practice what we have just learned, we will code a program that could be useful when grading the students from `STAT 294`. 

<div class = 'space'></div>

## Creating new classes

To do so, we will start by creating a class for the `students`.

```{r}
# Students
students <- function (names = character(), ...) {
  stopifnot(is.character(names))
  structure(list(names = names), class = 'students')
}

stds_names <- sort(c('ANA', 'HATOON', 'KARLA', 'MALIK', 'MARTINA', 'QILONG', 'RUIMAN', 'YASMEEN'))
stds <- students(names = stds_names)
stds
```

Notice that many other information about students could have been included here (e.g., `id`, `age`, etc.), but for this example, we are only interested in their `names`.

Now, let's create a class for the projects. In this case, I will create a separate function for validation, namely `validade_projects()`.

```{r}

# Projects
validate_projects <- function (students, prjs_title, grades, ...) {
  if (!('students' %in% class(students))) {
    stop('"students" should be an object of class "students".')
  } else if (!is.character(prjs_title)) {
    stop('"prjs_title" should be an object of class "character".')
  } else if (!is.numeric(grades)) {
    stop('"grades" should be an object of class "numeric".')
  } else if (!is.matrix(grades)) {
    stop('"grades" should be a matrix.')
  }
  
  names <- students$names
  n_std <- length(names)
  n_prj <- length(prjs_title)
  
  if (nrow(grades) != n_std) {
    stop(paste('"grades" matrix should have', n_std, 'rows.'))
  } else if (ncol(grades) != n_prj) {
    stop(paste('"grades" matrix should have', n_prj, 'cols.'))
  }
}

projects <- function(students, prjs_title = character(), grades = numeric(), ...) {
  validate_projects(students = students, prjs_title = prjs_title, grades = grades)
  structure(list(names = students$names,
                 projects = prjs_title,
                 grades = grades), class = 'projects')
}

prjs_title <- c('Dashboard', 'Shiny App', 'R package')
prjs_grade <- matrix(data = round(runif(n = (length(stds_names) * length(prjs_title)), min = 0, max = 10)), nrow = length(stds_names), byrow = FALSE)

prjs <- projects(students = stds, prjs_title = prjs_title, grades = prjs_grade)
prjs
```

As a remark, notice that the `students` attribute has to an object of class `students` (as we defined before). Also, for the `prjs` object, I have information about students' names, grades in each project (`projects`), and the projects themselves.

<div class = 'space'></div>

## Creating auxiliary functions

Now, suppose that we want to implement a function that will randomly select an order for students' projects presentations. To do so, we will create a function named `presentation_schedule()`.
```{r}
# Utils
presentation_schedule <- function (x, dates, presentations_per_day) {
  if (!(class(x) %in% c('students', 'projects'))) {
    stop('"x" should be an object of classes "students" or "projects".')
  } else if (length(x$names) > (length(dates) * presentations_per_day)) {
    stop('The number of total presentations is larger than the number of dates times the number of presentations per day.')
  }
  
  order_names <- sample(x = x$names, size = length(x$names), replace = FALSE)
  prest_sched <- matrix(data = NA, nrow = presentations_per_day, ncol = length(dates))
  
  count <- 1
  for (j in 1:ncol(prest_sched)) {
    for (i in 1:nrow(prest_sched)) {
      if (count <= length(order_names)) {
        prest_sched[i, j] <- order_names[count] 
      }
      count <- count + 1
    }
  }
  
  prest_sched <- as.data.frame(prest_sched)
  colnames(prest_sched) <- dates
  prest_sched
}

presentation_schedule(x = stds, dates = c('Monday', 'Thursday'), presentations_per_day = 5)
```

For the `presentation_schedule()` function, notice that we also require `x` to be an object of class `students` **or** `projects`.

<div class = 'space'></div>

## Extending generic functions

Now that we have objects from two new classes, namely `students` and `projects`, we can define new behaviors for **generic functions** (also, we could create our own generic functions). In particular, we will code new behaviors for the functions `print()`, `summary()`, and `plot()`.

<div class = 'space'></div>

### `print()`

Let's start by creating a method `print.projects()`. So, different from the standard output, we would like to print a data frame for the all grades with meaningful row and column names.
```{r}
# Before
print(prjs)

print.projects <- function (x, ...) {
  names <- x$names
  projects <- x$projects
  grades <- x$grades
  
  obj <- as.data.frame(grades)
  rownames(obj) <- names
  colnames(obj) <- projects
  
  print(obj)
  invisible(obj)
}

# After
print(prjs)
```

<div class = 'space'></div>

### `summary()`

Now, suppose that we would like to have quick access to a few summary statistics (e.g., `mean`, `sd`, `min`, `max`, etc.) of the students' grades. To do this, we can define a new behavior for `summary()` by creating the `summary.projects()` function.
```{r}
summary.projects <- function (object, ...) {
  projects <- object$projects
  grades <- object$grades
  
  for (i in 1:length(projects)) {
    cat(projects[i], '\n', sep = '')
    cat('mean: ', mean(grades[, i]), '\n', sep = '')
    cat('sd:   ',   sd(grades[, i]), '\n', sep = '')
    cat('min:  ',  min(grades[, i]), '\n', sep = '')
    cat('max:  ',  max(grades[, i]), '\n \n', sep = '')
  }
}

summary(prjs)
```

<div class = 'space'></div>

### `plot()`

Finally, let's create some customized plots. Here, we will add a different behavior for `plot()` function. But different from what we have just done for the `print()` and `summary()` functions, we will also allow different outputs for the same object depending on the value of an additional parameter (`type`). So, let's create our `plot.projects()` function.
```{r}
library(reshape2)

plot.projects <- function (x, type = 'boxplot', special_font = TRUE, ...) {
  names <- x$names
  projects <- x$projects
  grades <- x$grades
  
  obj <- as.data.frame(grades)
  obj['NAMES'] <- names
  rownames(obj) <- names
  colnames(obj) <- c(projects, 'NAMES')
  
  long_obj <- reshape2::melt(data = obj)
  
  if (type == 'boxplot') {
    if (special_font) { par(family = 'LM Roman 10', las = 1) } else { par(family = '', las = 1) }
    boxplot(formula = value ~ variable, data = long_obj, xlab = 'Projects', ylab = 'Grades', main = 'Distribution of grades for all projects')
  } else if (type == 'points') {
    if (special_font) { par(family = 'LM Roman 10', las = 2) } else { par(family = '', las = 2) }
    plot(NA, xlim = c(1, length(names)), ylim = c(0, (max(grades) * 1.25 )), xlab = '', ylab = 'Grades', main = 'Students\' grades for all projects', xaxt = 'n')
    axis(1, at = seq(1, length(names)), labels = names)
    for (i in 1:(length(projects))) {
      points(x = seq(1, length(names)), y = obj[ , i], col = (i + 1), pch = 19)
    }
    legend(x = 'topright', legend = projects, pch = rep(19, length(projects)), col = seq(2, (length(projects) + 1)))
  } else {
    stop('Select a valid "type".')
  }
}
```

As a comment, notice that we are internally using the `melt()` function from the `reshape2` package. To use our `plot.projects()` function, we can do as follows
```{r}
plot(x = prjs, type = 'points')
plot(x = prjs) # "type = 'boxplot'" is the default
```

<div class = 'space'></div>

## Complete program

The entire code for this section can be seen below.

```{r, eval = FALSE}
library(reshape2)

# Students
students <- function (names = character(), ...) {
  stopifnot(is.character(names))
  structure(list(names = names), class = 'students')
}

# Projects
validate_projects <- function (students, prjs_title, grades, ...) {
  if (!('students' %in% class(students))) {
    stop('"students" should be an object of class "students".')
  } else if (!is.character(prjs_title)) {
    stop('"prjs_title" should be an object of class "character".')
  } else if (!is.numeric(grades)) {
    stop('"grades" should be an object of class "numeric".')
  } else if (!is.matrix(grades)) {
    stop('"grades" should be a matrix.')
  }
  
  names <- students$names
  n_std <- length(names)
  n_prj <- length(prjs_title)
  
  if (nrow(grades) != n_std) {
    stop(paste('"grades" matrix should have', n_std, 'rows.'))
  } else if (ncol(grades) != n_prj) {
    stop(paste('"grades" matrix should have', n_prj, 'cols.'))
  }
}

projects <- function(students, prjs_title = character(), grades = numeric(), ...) {
  validate_projects(students = students, prjs_title = prjs_title, grades = grades)
  structure(list(names = students$names,
                 projects = prjs_title,
                 grades = grades), class = 'projects')
}

# Utils
presentation_schedule <- function (x, dates, presentations_per_day) {
  if (!(class(x) %in% c('students', 'projects'))) {
    stop('"x" should be an object of classes "students" or "projects".')
  } else if (length(x$names) > (length(dates) * presentations_per_day)) {
    stop('The number of total presentations is larger than the number of dates times the number of presentations per day.')
  }
  
  order_names <- sample(x = x$names, size = length(x$names), replace = FALSE)
  prest_sched <- matrix(data = NA, nrow = presentations_per_day, ncol = length(dates))
  
  count <- 1
  for (j in 1:ncol(prest_sched)) {
    for (i in 1:nrow(prest_sched)) {
      if (count <= length(order_names)) {
        prest_sched[i, j] <- order_names[count] 
      }
      count <- count + 1
    }
  }
  
  prest_sched <- as.data.frame(prest_sched)
  colnames(prest_sched) <- dates
  prest_sched
}


# Generic functions
print.projects <- function (x, ...) {
  names <- x$names
  projects <- x$projects
  grades <- x$grades
  
  obj <- as.data.frame(grades)
  rownames(obj) <- names
  colnames(obj) <- projects
  
  print(obj)
  invisible(obj)
}
# rm(print.projects)

summary.projects <- function (object, ...) {
  projects <- object$projects
  grades <- object$grades
  
  for (i in 1:length(projects)) {
    cat(projects[i], '\n', sep = '')
    cat('mean: ', mean(grades[, i]), '\n', sep = '')
    cat('sd:   ',   sd(grades[, i]), '\n', sep = '')
    cat('min:  ',  min(grades[, i]), '\n', sep = '')
    cat('max:  ',  max(grades[, i]), '\n \n', sep = '')
  }
}

plot.projects <- function (x, type = 'boxplot', special_font = TRUE, ...) {
  names <- x$names
  projects <- x$projects
  grades <- x$grades
  
  obj <- as.data.frame(grades)
  obj['NAMES'] <- names
  rownames(obj) <- names
  colnames(obj) <- c(projects, 'NAMES')
  
  long_obj <- melt(data = obj)
  
  if (type == 'boxplot') {
    if (special_font) { par(family = 'LM Roman 10') } else { par(family = '') }
    boxplot(formula = value ~ variable, data = long_obj, xlab = 'Projects', ylab = 'Grades', main = 'Distribution of grades for all projects')
  } else if (type == 'points') {
    if (special_font) { par(family = 'LM Roman 10') } else { par(family = '') }
    plot(NA, xlim = c(1, length(names)), ylim = c(0, (max(grades) * 1.25 )), xlab = 'Students', ylab = 'Grades', main = 'Students\' grades for all projects', xaxt = 'n')
    axis(1, at = seq(1, length(names)), labels = names)
    for (i in 1:(length(projects))) {
      points(x = seq(1, length(names)), y = obj[ , i], col = (i + 1), pch = 19)
    }
    legend(x = 'topright', legend = projects, pch = rep(19, length(projects)), col = seq(2, (length(projects) + 1)))
  } else {
    stop('Select a valid "type".')
  }
}

# Main code

stds_names <- sort(c('ANA', 'HATOON', 'KARLA', 'MALIK', 'MARTINA', 'QILONG', 'RUIMAN', 'YASMEEN'))
prjs_title <- c('Dashboard', 'Shiny App', 'R package')
prjs_grade <- matrix(data = round(runif(n = (length(stds_names) * length(prjs_title)), min = 0, max = 10)), nrow = length(stds_names), byrow = FALSE)

stds <- students(names = stds_names)
prjs <- projects(students = stds, prjs_title = prjs_title, grades = prjs_grade)

presentation_schedule(x = stds, dates = c('Monday', 'Thursday'), presentations_per_day = 5)

print(x = prjs)
summary(object = prjs)
plot(x = prjs, type = 'points')
plot(x = prjs, type = 'boxplot')
```

<div class = 'space'></div>

## Final comments

As a remark, notice that this small project (after some changes; for instance, by writing a proper documentation for all functions) could have been done as an `R` package. To see how to do this, refer to [this tutorial](rpack.html). 

